Alberta's Electricity Market

### 2022-01-24 paper/paper_draft_ws.tex

\documentclass[12pt]{article}
\usepackage{fullpage}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage[sort&compress]{natbib}
\usepackage{amssymb, amsmath, amsthm}
\usepackage[breaklinks=true]{hyperref}
\hypersetup{
    colorlinks,
    citecolor= blue,
    filecolor= blue,
    linkcolor= blue,
    urlcolor= blue
}

\usepackage{setspace}
\usepackage{tikz}
\usetikzlibrary{patterns,arrows,decorations.pathreplacing}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{multicol}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage[normalem]{ulem}

@@

\usepackage[hang,flushmargin]{footmisc}

@@

\urlstyle{same}

@@

%%%% TITLE %%%%
\title{Marginal and Average Carbon Price Pass-Through\thanks{We are grateful to Sam Harrison for excellent
research assistance. We thank seminar and conference participants at the AERE Summer Meetings for comments and suggestions. Financial support for this work was provided by the Future Energy Systems Canada First Research Excellence Fund grant held by the University of Alberta.}
}

@@

We evaluate the response of wholesale power market participants in Alberta to a series of changes in greenhouse gas (GHG) emissions policies.  Between 2015 and 2019, legislative and regulatory changes have affected carbon price levels, the rates of output-based allocations of emissions credits, and the coverage of carbon pricing across facilities of different sizes, creating variation in average and marginal carbon prices both across and within facility types and across and within portfolios held by major players in Alberta's wholesale power market. We exploit this variance in treatment to provide unique evidence of carbon price pass-through in power markets. We show that even after adjusting for costs imposed by carbon policies at the facility level, statistically significant changes in portfolio-level offer curves were induced leading to a shift in the merit order toward gas and away from coal which exceeded that which would have been predicted by carbon prices alone. These changes have accelerated the phase-out of coal-fired power in the province.

The Alberta wholesale power market provides an excellent quasi-laboratory because it is small, isolated, and the market is settled on a real-time, energy-only basis. As such, there are limited confounding factors beyond what is observable in our data. Alberta's market is winter-peaking, with peak 2021 load of just under 12,000 MW.\footnote{The 2021 maximum Alberta internal load per AESO 2021 data was 11,729 MW at time of writing, which is also the peak value ever recorded in Alberta.} Alberta has only three small interconnections to British Columbia, Montana, and Saskatchewan, with total capacity of less than 15\% of peak internal load.\footnote{The British Columbia and Montana tie lines have a joint operating capacity of 1500 MW for imports and 1325 MW for export, while the Saskatchewan tie line has a path rating of 150 MW.} The relative isolation of Alberta's market means that local prices and generation will be more affected by Alberta's policies than would be the case if traded power had a more important market share, and implies that extra-jurisdictional confounding factors are less likely to influence our results. The downside to this is that Alberta, a major oil-producing jurisdiction, was hit with a major, oil-price-induced recession contemporaneously with some of the changes in carbon policies, and this economic downturn led to a period of low prices in the power market. Alberta's generation mix is also somewhat unique, with a significant portion of internal load served by combined heat and power (or co-generation) plants associated with industrial facilities, primarily located in the oil sands region in Northern Alberta. Despite these anomalies, the transparency and isolation of Alberta's market provide us with a unique opportunity to study carbon price pass-through to prices as well as the induced changes in market participation.

We compile over 10 years of hourly data describing market offer behaviour and a wide range of relevant characteristics. We measure and describe facility-level hourly offers of power into the market for the time period preceding the COVID-19 pandemic. We combine these data with detailed plant characteristics compiled from regulatory data, emissions compliance data from Alberta's and Canada's air emissions reporting, weather data, commodity price information, and power market data including hourly renewable generation, imports and exports, import and export capabilities and total and forecast internal loads. Finally, we include both observed carbon policy costs and the value of output-based allocations at the facility level.

In order to isolate the impact of the changes in policies on offer behaviour, we develop a unique empirical strategy...

@@

The paper proceeds as follows. In Section \ref{sec:policy}, we detail Alberta's greenhouse gas policy history. In Sections \ref{sec:data} and \ref{sec:market} we detail our data set and provide a statistical portrait of the province's power market. In Section \ref{sec:empirical} we detail our empirical strategy. Section \ref{sec:results} summarizes our results. Section \ref{sec:conclusion} concludes.

\section{Alberta's GHG Policy Changes}\label{sec:policy}

Alberta has had carbon pricing in place since the \emph{Specified Gas Emitters Regulation} (\emph{SGER}) took effect on July 1, 2007.\footnote{\textit{Specified Gas Emitters Regulation}, Alta Reg 139/2007, $<$\url{canlii.ca/t/52x2q}$>$.} That regulation, the first industrial carbon price in North America, implemented a price of \$15/tonne, and allocated emissions credits to covered facilities at a rate equal to 88\% of an individual facility's historic (2003-2005) emissions intensity. For new facilities built after 2005, they were allocated emissions credits per unit output at a rate equal to 88\% of their average year 3 emissions intensity. In addition to carbon pricing on industrial emissions, the \emph{SGER} also included an offset protocol which provided emissions credits for deemed emissions reductions due to certain activities. In the case of combined heat and power plants, facilities received an allocation of credits which amounted to 0.418t/MWh for net-to-grid electricity. New renewable power facilities were also eligible for offset credits under the SGER, with a deemed emissions reduction rate of 0.65t/MWh.\footnote{Cite. Subsequently reduced to 0.59t/MWh (2015) and 0.53t/MWh (2019)}

This system remained in place until June, 2015 when the government of Premier Rachel Notley introduced a series of changes to the existing regulation.\footnote{\textit{Specified Gas Emitters Amendment Regulation}, Alta Reg 105/2015, $<$\url{perma.cc/Z2ZM-WDWP}$>$.} The first set of changes increased the carbon price to \$20/tonne for 2016 and to \$30/tonne for 2017, while also reducing the benchmarks for the output-based allocation of emissions credits to 85\% and 80\% of historic facility level emissions intensity for 2016 and 2017 respectively. Combined, these changes implied a material increase in the average cost of carbon in each of the years 2016 and 2017. The government subsequently adopted, in November of 2015, a more comprehensive change to GHG emissions policies. Two changes in this iteration of policies affected power markets.  Most importantly, the \textit{Carbon Competitiveness Incentive Regulation (CCIR)} replaced the \emph{SGER} and these regulations levelized the output-based allocation of emissions credits across all power generators at 0.37t/MWh, the emissions intensity of the best-in-class combined cycle natural gas generation facility in the province.\footnote{\textit{Carbon Competitiveness Incentive Regulation}, Alta Reg 255/2017, $<$\url{canlii.ca/t/53h9v}$>$.} This implied that coal producers saw a steep increase in their average costs of carbon, while impacts on gas power plants varied depending on the heat rates of the facility. Both combined heat and power plants and existing renewable generation facilities saw the value of their emissions credits issued per MWh generated decrease under the \emph{CCIR}. The second important change was that an economy-wide carbon price was introduced for facilities not covered under the \emph{CCIR} - those without historic emissions in any previous year greater than 100,000 tonnes. These facilities did not, by default, receive output-based allocations of emissions credits to offset the cost of the carbon price, so their average costs of carbon could be much higher than their larger competitors.  An opt-in provision allowing these smaller firms to be covered under the \emph{CCIR} was available, should facilities wish to undertake the more comprehensive emissions reporting. While not directly material to our sample period, the government also announced an accelerated coal phase-out which would see all coal-fired generation shut-down, equipped with carbon capture and storage, or re-fired with natural gas by December 31, 2030.

The \emph{CCIR} remained in place until it was replaced by the \emph{
Technology Innovation and Emissions Reduction Regulation} (\emph{TIER}) which took effect on January 1, 2020, outside of our sample period.\footnote{\textit{Technology Innovation and Emissions Reduction Regulation}, Alta Reg 133/2019, $<$\url{canlii.ca/t/54qgg}$>$.}  However, it's important to note that while \emph{TIER} altered the design of output-based allocations in most sectors, it did not impose material changes in the electricity industry and so we should not expect any changes late in our sample period which abuts the implementation of \emph{TIER}. \emph{TIER} remains in place as of this writing, with the price on carbon having increased to \$40 per tonne for 2021 and \$50 per tonne for 2022.\footnote{See Alberta Environment and Parks Ministerial Order 87-2021 under the \emph{Emissions Management and Climate Resilience Act}, SA 2003, c. E-7.8, $<$\url{perma.cc/3HN3-H9RC}$>$.}

>>>>

\documentclass[12pt]{article}
\usepackage{fullpage}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage[sort&compress]{natbib}
\usepackage{amssymb, amsmath, amsthm}
\usepackage[breaklinks=true]{hyperref}
\hypersetup{
    colorlinks,
    citecolor= black,
    filecolor= black,
    linkcolor= black,
    urlcolor= black
}
\usepackage{setspace}
\usepackage{tikz}
\usetikzlibrary{patterns,arrows,decorations.pathreplacing}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{multicol}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage[normalem]{ulem}

@@

\usepackage[multiple]{footmisc}

@@

@@

%%%% TITLE %%%%
\title{Carbon Price Pass-Through in Alberta's Electricity Market\thanks{We are grateful to Sam Harrison for excellent
research assistance. We thank seminar and conference participants at the 2019 AERE Summer Meetings for comments and suggestions. Financial support for this work was provided by the Future Energy Systems Canada First Research Excellence Fund grant.
}
}

@@

We evaluate the response of wholesale power market participants in Alberta to a series of changes in greenhouse gas (GHG) emissions policies.  Between 2015 and 2019, legislative changes have affected carbon price levels, the rates of output-based allocations of emissions credits, and the coverage of carbon pricing across facilities of different sizes, creating variation in average and marginal carbon prices both across and within facility types and across and within portfolios held by major players in Alberta's wholesale power market. We exploit this variance in treatment to provide unique evidence of offer response to carbon pricing in power markets. We show that even after adjusting for costs imposed by carbon policies at the facility level, statistically significant changes in portfolio-level offer curves were induced leading to a shift in the merit order toward gas and away from coal which exceeded that which would have been predicted by carbon prices alone.

The Alberta wholesale power market provides an excellent laboratory for our study because it is small, isolated, and the market is settled on a real-time, energy-only basis, so there are limited confounding factors beyond what is observable in our data. Alberta's market is winter-peaking, with peak 2021 load of just under 12,000 MW.\footnote{The 2021 maximum Alberta internal load per AESO 2021 data was 11,729 MW at time of writing, which is also the peak value ever recorded in Alberta.} Alberta has only three small interconnections to British Columbia, Montana, and Saskatchewan, with total capacity of less than 15\% of peak internal load.\footnote{The British Columbia and Montana tie lines have a joint capacity of 1500 MW for imports and 1325 MW for export, while the Saskatchewan tie line has a path rating of 150 MW.} The relative isolation of Alberta's market means that local prices and generation will be more affected by Alberta's policies than would be the case if traded power had a more important market share, and implies that extra-jurisdictional confounding factors are less likely to influence our results. The downside to this is that Alberta, a major oil-producing jurisdiction, was hit with a major, oil-price-induced recession contemporaneously with some of the changes in carbon policies, and this economic downturn led to a period of low prices in the power market. Alberta's generation mix is also somewhat unique, with a significant portion of internal load served by combined heat and power (or co-generation) plants associated with industrial facilities, primarily located in the oil sands region in Northern Alberta. Despite these anomalies, the transparency and isolation of Alberta's market provide us with a unique opportunity to study carbon price pass-through to prices as well as the induced changes in market participation.

We compile data describing market offer behaviour and a wide range of relevant covariates. We measure and describe facility-level hourly offers of power into the market over slightly more than 10 years preceding the COVID-19 pandemic. We combine these data with detailed plant characteristics compiled from regulatory data, emissions compliance data from Alberta's and Canada's air emissions reporting, weather data, commodity price information, and power market data including hourly renewable generation, imports and exports, import and export capabilities and total and forecast internal loads. Finally, we include both observed carbon policy costs and the value of output-based allocations at the facility level.

In order to isolate the impact of the changes in policies on offer behaviour, we develop a unique empirical strategy based on synthetic power plants and synthetic power portfolios.  For each hour of each day, we create synthetic power portfolios either by plant type .  We are then able to examine, for example, how changes in policies affected the offer of power across all coal- or gas-fired plants in the province.

or by which entity holds offer control on the units in question
, or across all facilities owned by particular actors in the market.  We normalize these portfolios by percentile, so we are able to compare behaviour across portfolios of different sizes without loss of generality.

@@

\section{Alberta's GHG Policy Changes}

Alberta has had carbon pricing in place since the \emph{Specified Gas Emitters Regulation} (\emph{SGER}) took effect on July 1, 2007. That regulation, the first industrial carbon price in North America, implemented a price of \$15/tonne, and allocated emissions credits to covered facilities at a rate equal to 88\% of an individual facility's historic (2003-2005) emissions intensity. For new facilities built after 2005, they were allocated emissions credits per unit output at a rate equal to 88\% of their average year 3 emissions intensity. In addition to carbon pricing on industrial emissions, the \emph{SGER} also included an offset protocol which provided emissions credits for deemed emissions reductions due to certain activities. In the case of combined heat and power plants, facilities received an allocation of credits which amounted to 0.418t/MWh for net-to-grid electricity. New renewable power facilities were also eligible for offset credits under the SGER, with a deemed emissions reduction rate of 0.65t/MWh.\footnote{Cite. Subsequently reduced to 0.59t/MWh (2015) and 0.53t/MWh (2019)}

This system remained in place until June, 2015 when the government of Premier Rachel Notley introduced a series of changes to the existing regulation. The first set of changes increased the carbon price to \$20/tonne for 2016 and to \$30/tonne for 2017, while also reducing the benchmarks for the output-based allocation of emissions credits to 85\% and 80\% of historic facility level emissions intensity for 2016 and 2017 respectively. Combined, these changes implied a material increase in the average cost of carbon in each of the years 2016 and 2017. The government subsequently adopted, in November of 2015, a more comprehensive change to GHG emissions policies. Two changes in this iteration of policies affected power markets.  Most importantly, the \textit{Carbon Competitiveness Incentive Regulation (CCIR)} replaced the \emph{SGER} and these regulations levelized the output-based allocation of emissions credits across all power generators at 0.37t/MWh, the emissions intensity of the best-in-class combined cycle natural gas generation facility in the province. This implied that coal producers saw a steep increase in their average costs of carbon, while impacts on gas power plants varied depending on the heat rates of the facility. Both combined heat and power plants and existing renewable generation facilities saw the value of their emissions credits issued per MWh generated decrease under the \emph{CCIR}. The second important change was that an economy-wide carbon price was introduced for facilities not covered under the \emph{CCIR} - those without historic emissions in any previous year greater than 100,000 tonnes. These facilities did not, by default, receive output-based allocations of emissions credits to offset the cost of the carbon price, so their average costs of carbon could be much higher than their larger competitors.  An opt-in provision allowing these smaller firms to be covered under the \emph{CCIR} was available, should facilities wish to undertake the more comprehensive emissions reporting. While not directly material to our sample period, the government also announced an accelerated coal phase-out which would see all coal-fired generation shut-down, equipped with carbon capture and storage, or re-fired with natural gas by December 31, 2030.

The \emph{CCIR} remained in place until it was replaced by the \emph{
Technology Innovation and Emissions Reduction Regulation} (\emph{TIER}) on January 1, 2020, outside of our sample period. However, it's important to not that while \emph{TIER} altered the design of output-based allocations in most sectors, it did not impose material changes in the electricity industry and so we should not expect any changes late in our sample period which abuts the implementation of \emph{TIER}.
